Typing a linear \picalc

### 2020-04-27 paper/paper.tex
Comment on subject reduction

% Functions
\newcommand{\subst}[3]{#1[#2/#3]}
\newcommand{\opsquared}[3]{#1 \, \coloneqq \, #2 \, \cdot^2 \, #3}
\newcommand{\opctx}[3]{#1 \, \coloneqq \, #2 \, \otimes \, #3}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}

@@

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common variable.
The receiving process substitutes references to its most immediate variable with references to the variable sent by the process doing the output, and is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- but notably not under input nor output, otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the variable over which communication happens.
If this variable is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the variable index over which communication occurs to the type level.
Every time we come out of a binder we decrement this variable.
To do so, we make use of a $dec$ function that saturates at $\constr{internal}$ --- as opposed to an $inc$ function that would not.

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n \\ uP : \Unused_0 \, P}
  {\constr{comm} : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{lower \; 0 \; \subst{P}{j}{0} \; uP}{Q}}

@@

Subject reduction (aka type preservation) relates the operational semantics and the typing system of a language.
In its usual form, it states that if $P$ reduces to $Q$ and $P$ is well-typed under context $\Gamma$, then $Q$ is well-typed under $\Gamma$ too.
In the linear \picalc{} however, reduction \emph{is} meant to transform the context: each time communication occurs over a channel, that channel is exhausted.
In our case, and because we generalize linearity to a partial monoid, the channel over which the communication occurs will loose a multiplicity on both its input and output capabilities.
In \S \ref{semantics}, we made sure to lift the index of this channel to the type level.
If the channel is internal to the process, the context must not be modified; if it is external on index $i$, the input usage context at index $i$ must have $\ell_\#$ subtracted.

\begin{nitheorem}
  Let $P \reduce{c} Q$.
  If $c$ is $\constr{internal}$, then let $\Delta \equiv \epsilon$;
  if $c$ is $\constr{external}\; i$, then let $\Delta \equiv \ell_\# @ i$.
  Let $\Gamma'$ and $\Gamma$ be usage contexts only differing in $\Delta$, so that $\opctx{\Gamma'}{\Delta}{\Gamma}$.
  If $\types{\gamma}{\Gamma'}{P}{\Xi}$ then $\types{\gamma}{\Gamma}{Q}{\Xi}$
\end{nitheorem}

\begin{nilemma}
  \label{lm:comm-capable}
  \todo{reformulate}
  A well-typed process on which reduction occurs externally, must have this external multiplicity be capable of $\ell_\#$.
\end{nilemma}

\begin{proof}
  For the base case $\constr{comm}$ \todo{}

  Reduction under composition with $\constr{par}$ \todo{}

  Reduction under scope restriction case splits on \textsc{Channel} $c$ on which communication occurs.
  If this channel is internal to the process, or if it is external but on a channel other then the one introduced by the recent scope restriction, subject reduction proceeds inductively.
  However, if reduction happens on the most recently introduced channel, we use the Lemma \ref{lm:comm-capable} to subtract $\ell_\#$ from the channel's multiplicity, then apply the induction hypothesis.

  Finally, in the $\constr{struct}$ case, we apply subject congruence (Theorem \ref{thm:subject-congruence}) and then the induction hypothesis.

\end{proof}

@@

\begin{nitheorem}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per Lemma \ref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let us insert an arbitrary type $t$ at position $i$ in the typing context and an arbitrary multiplicity $x$ at position $i$ into both the input and output contexts.
  Let us reflect this on the process $P$ by incrementing every index greater than or equal to $i$.
  The resulting process will be well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index that is not mentioned by $P$, such that $\Unused \; i \; P$.
  Then removing index $i$ from the typing context and both of the usage contexts and decrementing every variable in $P$ greater than $i$ preserves its well typedness, so that $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  If $\types{\gamma}{\Gamma}{P}{\Xi}$ then $\types{swap_i \; \gamma}{swap_i \; \Gamma}{swap_i \; P}{swap_i \; \Xi}$.
\end{nitheorem}

@@

\begin{nitheorem}
  Let $P$ be a process where the innermost variable has all its $m$ multiplities exhausted, such that $\types{\gamma , t}{\Gamma , m}{P}{\Psi , \ell_{\o}}$.
  Let $j$ be the index of the data that is being received.
  This data has at least $m$ multiplicities, such that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the innermost variable by $j$.
  As a result, we will exhaust no multiplicities on the innermost variable and exhaust them in $j$ instead, such that $\types{\gamma , t}{\Gamma , m}{\subst{P}{j}{\constr{zero}}}{\Xi , m}$.
\end{nitheorem}

\begin{proof}
  By induction on the typing derivation on $P$.
  To be able to proceed under scope restriction, we need to generalize the theorem so that the variable that is being substituted need not be $\constr{zero}$, but any variable $i$.
  We introduce such generalization in Theorem \ref{th:subst-generalization}.
  \todo{explain arguments}
\end{proof}

\begin{nitheorem}
  \label{th:subst-generalization}

>>>>

% Functions
\newcommand{\subst}[3]{#1 \; [ \; #2 \; / \;#3 \;]}
\newcommand{\opsquared}[3]{#1 \, \coloneqq \, #2 \, \cdot^2 \, #3}
\newcommand{\opctx}[3]{#1 \, \coloneqq \, #2 \, \otimes \, #3}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{\propto} \; #2 \; \type{\vdash} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\contains}[6]{#1 \; \type{\propto} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{\propto} \; #5 \; \type{\boxtimes} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\boxtimes} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Carrier}{\type{CARRIER}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Quantifier}{\type{QUANTIFIER}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}

@@

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common variable.
The receiving process substitutes references to its most immediate variable with references to the variable sent by the process doing the output.
The process is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- but notably not under input nor output, otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the variable over which communication happens.
If this variable is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the variable index over which communication occurs to the type level.
Every time we come out of a binder we decrement this variable.
To do so, we make use of a $dec$ function that saturates at $\constr{internal}$ --- as opposed to an $inc$ function that would not.

@@

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n}
  {\constr{comm} : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce{\constr{external} \; i} \comp{lower \; 0 \; (\subst{P}{j}{0})}{Q}}

@@

Subject reduction (aka type preservation) relates the operational semantics and the typing system of a language.
In its usual form, it states that if $P$ reduces to $Q$ and $P$ is well-typed under context $\Gamma$, then $Q$ is well-typed under $\Gamma$ too.
In the linear \picalc{} however, reduction \emph{is} meant to transform the context: each time communication occurs over a channel, that channel is exhausted.
In our case, and because we generalize linearity to a partial monoid, the channel over which the communication occurs will loose a multiplicity $\one$ on both its input and output capabilities.
In \S \ref{semantics}, we made sure to lift the index of the communicating channel to the type level, now we use it to state subject reduction.

\begin{nitheorem}
  Let $P \reduce{c} Q$.
  If $c$ is $\constr{internal}$ to the process, the context must not be modified and thus $\Gamma' \equiv \Gamma$.
  If $c$ is $\constr{external}\; i$, the input usage context at index $i$ must be decremented, and therefore $\containsusage{\Gamma'}{i}{\ell_\#}{\Gamma}$.
  Then the typing derivation $\types{\gamma}{\Gamma'}{P}{\Xi}$ can be transformed into $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the base case $\constr{comm}$ we need to construct a typing derivation $\types{\gamma}{\Gamma}{lower \; 0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi}$ from the following bits:
    \begin{alignat*}{2}
      \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
      \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
      \types{\gamma , t}{&\Psi , m}{P &&}{\Delta , \ell_{\o}} \\
      \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
      \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
    \end{alignat*}

    We start by using framing (Theorem \ref{thm:framing}) to rearrange the consumption of the input and output multiplicities of the communicating channel, so that we obtain the following:
    \begin{alignat*}{2}
      \types{\gamma , t}{& \Gamma , m}{P &&}{\Theta , \ell_{\o}} \\
      \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
    \end{alignat*}

    We then use substitution (Theorem \ref{thm:substitution}) to obtain the following:
    \begin{align*}
      \types{\gamma , t}{\Gamma , m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi , m}
    \end{align*}

    Finally, we use strengthening (Theorem \ref{thm:strengthening}) to lower the substituted variable index:
    \begin{align*}
      \types{\gamma , t}{\Gamma , m}{lower \; 0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi , m}
    \end{align*}

    \item
    Reduction under composition with $\constr{par}$ proceeds by induction on the process that is being reduced.

    \item
    Reduction under scope restriction case splits on \textsc{Channel} $c$ on which communication occurs.
    If this channel is internal to the process, or if it is external but on a channel other then the one introduced by the recent scope restriction, subject reduction proceeds inductively.
    However, if reduction happens on channel that is being introduced, we use the Lemma \ref{lm:comm-capable} to subtract $\ell_\#$ from the channel's multiplicity first, then apply the induction hypothesis.

    \item
    Finally, in the $\constr{struct}$ case, we apply subject congruence (Theorem \ref{thm:subject-congruence}) and then the induction hypothesis.
  \end{itemize}
\end{proof}

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ on a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces such that $P \reduce{\constr{external} \; i} Q$ has a multiplicity of at least $\ell_\#$ at index $i$.
\end{nilemma}
\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}

@@

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per Lemma \ref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  \label{thm:weakening}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let us insert an arbitrary type $t$ at position $i$ in the typing context and an arbitrary multiplicity $x$ at position $i$ into both the input and output contexts.
  Let us reflect this on the process $P$ by incrementing every index greater than or equal to $i$.
  The resulting process will be well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  \label{thm:strengthening}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index that is not mentioned by $P$, such that $\Unused \; i \; P$.
  Then removing index $i$ from the typing context and both of the usage contexts and decrementing every variable in $P$ greater than $i$ preserves its well typedness, so that $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}
  \label{thm:swapping}
  If $\types{\gamma}{\Gamma}{P}{\Xi}$ then $\types{swap_i \; \gamma}{swap_i \; \Gamma}{swap_i \; P}{swap_i \; \Xi}$.
\end{nitheorem}

\todo{}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the innermost variable has all its $m$ multiplities exhausted, such that $\types{\gamma , t}{\Gamma , m}{P}{\Psi , \ell_{\o}}$.
  Let $j$ be the index of the data that is being received.
  This data has at least $m$ multiplicities, such that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the innermost variable by $j$.
  As a result, we will exhaust no multiplicities on the innermost variable and exhaust them in $j$ instead, such that $\types{\gamma , t}{\Gamma , m}{\subst{P}{j}{\constr{zero}}}{\Xi , m}$.
\end{nitheorem}

\begin{proof}
  By induction on the typing derivation on $P$.
  To be able to proceed under scope restriction, we need to generalize the theorem so that the variable that is being substituted need not be $\constr{zero}$, but any variable $i$.
  We introduce such generalization in Theorem \ref{thm:subst-generalization}.
  \todo{explain arguments}
\end{proof}

\begin{nitheorem}
  \label{thm:subst-generalization}
